---
title: "Case Study: Daily Vitamin Use During Cold and Flu Season"
subtitle: "Learning Objective: Relative Risk and Odds Ratio"
date: "Last updated: `r Sys.Date()`"
output: 
  learnr::tutorial:
    theme: cosmo
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

## Introduction

Most people will tell you that Vitamin C is a good way to prevent getting sick during cold and flu season. But how do we know?

In 1972, a study was published in the *Canadian Medical Association Journal* titled "Vitamin C and the Common Cold". That study used a randomized experiment to assign 407 volunteers to consume 1,000 mg of vitamin C daily throughout the cold season, and 411 to receive a placebo. A physician interviewed the volunteers at the end of the study to determine whether or not they had suffered any colds during the study period.

```{r}
library(Sleuth3)
library(tidyverse)

data(case1802)

vitaminC <- case1802
```

The results of the study are summarized in the table below.

| Treatment | Developed a cold | Did not develop a cold | Total |
|-----------|------------------|------------------------|-------|
| Placebo   | 335              | 76                     | 411   |
| Vitamin C | 302              | 105                    | 407   |
| Total     | 637              | 181                    | 818   |

### Check your understanding

```{r quiz}
question("Why is randomization important in an experiment?",
  answer("Randomization ensures all treatment groups are the same size."),
  answer("Randomization ensures all participants have an even chance of receiving the treatment."),
  answer('Randomization ensures that any possible confounding variables are "balanced out".', correct=TRUE),
  answer("Randomization prevents possible researcher bias from affecting the results.", correct=TRUE)
)
```

Use the calculation box to find the proportion of all participants who developed a cold during the experiment.

```{r exercise1, exercise=TRUE}

```

```{r, exercise1-solution}
637/818
```

How does the likelihood of developing a cold compare between the vitamin C (treatment) and placebo group?

```{r}
data <- tribble(
  ~Group, ~Cold, ~n,
  "Placebo", "Yes", 335,
  "Placebo", "No", 76,
  "Vitamin C", "Yes", 302,
  "Vitamin C", "No", 105
)

data <- uncount(data, weights=n)

data %>%
  ggplot(aes(x=Group)) + 
  geom_bar(aes(fill=Cold), position='dodge')
```

## Sample proportions

One natural place to start is with the __sample proportions__.

$$\hat{p} = \frac{Number \ in \ category \ of \ interest}{Number \ in \ group}$$

Let $\hat{p}_1$ denote the sample proportion in the placebo group and $\hat{p}_2$ denote the sample proportion in the treatment group.


| Treatment | Developed a cold | Did not develop a cold | Total |
|-----------|------------------|------------------------|-------|
| Placebo   | 335              | 76                     | 411   |
| Vitamin C | 302              | 105                    | 407   |
| Total     | 637              | 181                    | 818   |

Let's begin with the placebo group. Out of 411 participants in the vitamin C group, 335 developed a cold.

$$\hat{p}_1 = \frac{335}{411} = 0.815 = 81.5 \%$$

In this experiment, 81.5% of participants in the placebo group developed a cold.

Use the calculation box below to find the sample proportion for the treatment group.

```{r exercise2, exercise=TRUE}

```

```{r, exercise2-solution}
302/407
```

```{r quiz2}
question("Which group was more likely to develop a cold during the experiment?",
  answer("Treatment group"),
  answer("Placebo group", correct=TRUE)
)
```

## Relative risk 

Proportions can be difficult to directly compare. The possible values of a proportion range from 0 to 1, which means that any difference between the proportions themselves will also be small.

### Relative risk

Relative risk is a... relative measure. 

> __Relative risk__: The ratio of the sample proportions of two groups

In other words, how large is the likelihood of something occurring in one group as a function of the other?




### Calculating relative risk

To calculate relative risk, find the sample proportions in each group first. Then, calculate the ratio of group 1 to group 2.

$$Relative \ risk = \frac{{\hat{p}}_1}{\hat{p}_2}$$

Use the calculation box below to find the relative risk of developing a cold for the placebo group compared to the treatment group.

| Treatment | Developed a cold | Did not develop a cold | Total |
|-----------|------------------|------------------------|-------|
| Placebo   | 335              | 76                     | 411   |
| Vitamin C | 302              | 105                    | 407   |
| Total     | 637              | 181                    | 818   |

```{r exercise3, exercise=TRUE}

```

```{r, exercise3-solution}
(335/411)/(302/407)
```

### Interpretation

Relative risk should be interpreted as a direct comparison.

```{r quiz3}
question("Choose the best interpretation of the relative risk.",
  answer("The risk of developing a cold while taking a placebo is 1.1 times the risk of developing a cold while taking vitamin C.", correct=TRUE),
  answer("The risk of developing a cold while taking vitamin C is 1.1 times the risk of developing a cold while taking a placebo."),
  answer("The risk of developing a cold while taking vitamin C is 1.1 times higher than the risk of developing a cold while taking a placebo."),
  answer("The risk of developing a cold while taking a placebo is 1.1 times higher than the risk of developing a cold while taking vitamin C.")
)
```

A relative risk close to 1 indicates that the two groups have similar proportions, or _risks_. The larger the relative risk, the greater the discrepancy between the risk of an event occurring in each group.

## Odds ratio

Odds ratio is another measure that can be used to compare the risk or probability of an event in two groups.

> __Odds ratio__: The ratio of the sample odds of two groups

In order to calculate the odds ratio, we first need to define the __odds__.

### Odds

> __Odds__: The ratio of "successes" (the category of interest) to "failures" (not the category of interest)

We can convert the sample proportion to the odds.

$$Odds = \frac{\hat{p}}{1-\hat{p}}$$

We previously showed that the proportion of all participants in the placebo group who developed a cold is $335/411 = 0.815$.

We can use this to find the odds for the placebo group.

$$Odds_1 = \frac{\hat{p}_1}{1-\hat{p}_1} = \frac{0.815}{1-0.815}=4.4$$

The __odds of developing a cold for the placebo group is 4.4__ - participants in this experiment were 4.4 times as likely to develop a cold as not. 


| Treatment | Developed a cold | Did not develop a cold | Total |
|-----------|------------------|------------------------|-------|
| Placebo   | 335              | 76                     | 411   |
| Vitamin C | 302              | 105                    | 407   |
| Total     | 637              | 181                    | 818   |

Use the calculation box below to find the odds for the treatment group.

```{r exercise4, exercise=TRUE}

```

```{r, exercise4-solution}
0.742/(1-0.742)

302/(105)
```


Now, we can revisit the odds ratio.

### Calculating odds ratio

To calculate odds ratio, find the sample proportions and/or in each group first. Then, calculate the ratio of the odds of group 1 to group 2.

$$Odds \ ratio =\frac{Odds_1}{Odds_2}= \frac{{\hat{p}}_1/(1-\hat{p}_1)}{\hat{p}_2/(1-\hat{p}_2)}$$

Use the calculation box below to find the odds of developing a cold for the placebo group compared to the treatment group.

_Hint_: $Odds_1=4.4$ and $Odds_2=2.9$.

```{r exercise5, exercise=TRUE}

```

```{r, exercise5-solution}
4.4/2.9
```

### Interpretation

Like relative risk, the odds ratio should be interpreted as a direct comparison.

```{r quiz4}
question("Choose the best interpretation of the odds ratio.",
  answer("The odds of developing a cold while taking vitamin C are 1.5 times the odds of developing a cold while taking a placebo."),
  answer("The odds of developing a cold while taking a placebo are 1.5 times the odds of developing a cold while taking vitamin C.", correct=TRUE)
)
```


## Practice

A five-year experiment was conducted to evaluate the effectiveness of fish oils on reducing cardiovascular events, where each subject was randomized into one of two treatment groups. 

Group|Heart attack|No heart attack|Total
----|----|----|----
Fish oil|145|12788|12933
Placebo|300|12638|12938

Use the calculation box to find the proportion of participants in the fish oil group who experienced a cardiac event.

```{r exercise6, exercise=TRUE}

```

```{r, exercise6-solution}
145/12933
```

Use the calculation box to find the proportion of participants in the placebo group who experienced a cardiac event.

```{r exercise7, exercise=TRUE}

```

```{r, exercise7-solution}
300/12938
```

Use the calculation box to find the relative risk of experiencing a cardiac event for the placebo group (1) compared to the fish oil group (2).

```{r exercise8, exercise=TRUE}

```

```{r, exercise8-solution}
0.0232/0.0112
```

Use the calculation box to find the odds ratio of experiencing a cardiac event for the placebo group (1) compared to the fish oil group (2).

```{r exercise9, exercise=TRUE}

```

```{r, exercise9-solution}
(0.0232/(1-0.0232))/(0.0112/(1-0.0112))
```


```{r quiz5}
question("Choose the best statement comparing these two groups in this experiment.",
  answer("Subjects who took a placebo supplement were about twice as likely to experience a cardiac event as subjects who took a fish oil supplement.", correct=TRUE),
  answer("Subjects who took a fish oil supplement were about twice as likely to experience a cardiac event as subjects who took a placebo supplement."),
  answer("There was no meaningful difference between the two experimental groups, since the likelihood of a cardiac event was very small in both.")
)
```