---
title: "Case Study: Breast Cancer Screenings"
subtitle: "Learning Objective: Conditional Probability and Bayes' Rule"
date: "Last updated: `r Sys.Date()`"
output: 
  learnr::tutorial:
    theme: cosmo
    progressive: true
    allow_skip: false
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

## Introduction

In Canada, it's estimated that about 0.35% of women over 40 will develop breast cancer in any given year. A common screening test for breast cancer is the mammogram, but like any diagnostic procedure it's not perfect. Suppose that about 7.3% of all mammograms will detect breast cancer.

Data suggests that in 11% of patients with breast cancer, the test gives a false negative. That is, the test indicates that a woman does not have breast cancer when in fact she does have it.

Similarly, the test gives a false positive in 7% of patients who do not have breast cancer. In other words it indicates these patients have breast cancer when they actually do not.

### Check your understanding

Let $B$ represent the event that a randomly selected Canadian woman over 40 develops breast cancer.

```{r quiz1}
question("Identify P(B), the probability that a randomly selected Canadian woman over 40 develops breast cancer.",
  answer("0.11"),
  answer("0.07"),
  answer("0.0035", correct=TRUE),
  answer("0.35")
)
```

## Conditional probability

One useful tool for representing and calculating probabilities is conditional probability.

> **Conditional probability**: The probability that an event $A$ occurs, given or conditional on the occurrence of $B$

If we know that event $B$ has occurred, does that change the probability that $A$ also occurs?

We denote the conditional probability of $A$ given $B$ as:

$$
P(A\vert B)
$$

The vertical bar reads as "given" or "conditional on".

In our breast cancer example, let $B$ denote the event a Canadian woman has breast cancer and $N$ denote the event a Canadian woman does not have breast cancer.

Further, let $+$ denote the probability that cancer is detected on a mammogram (positive result) and $-$ denote the probability that cancer is not detected (negative result).

We know that:

-   In about 11% of patients with breast cancer, the test gives a false negative. That is, the test indicates that a woman does not have breast cancer when in fact she does have it.

```{r quiz2}
question("What does this represent in probability notation?",
  answer("P(B|-)"),
  answer("P(-|B)", correct=TRUE),
  answer("P(B|+)"),
  answer("P(+|B)")
)
```

We also know that:

-   The test gives a false positive in 7% of patients who do not have breast cancer. In other words it indicates these patients have breast cancer when they actually do not.

```{r quiz4}
question("What does this represent in probability notation?",
  answer("P(N|-)"),
  answer("P(-|N)"),
  answer("P(N|+)"),
  answer("P(+|N)", correct=TRUE)
)
```

### Calculating conditional probability

To calculate a conditional probability, use the formula below.

$$P(A\vert B) = \frac{P(A \ and \ B)}{P(B)}$$

For our problem, we know:

- $P(B) = 0.0035$
- $P(-\vert B) = 0.11$
- $P(+ \vert N) = 0.07$

Suppose we want to find $P(- \ and \ B)$, the probability that a randomly selected Canadian woman both tests negative and truly has breast cancer.

$$P(-\vert B) = \frac{P(- \ and \ B)}{P(B)}$$

$$0.11 = \frac{P(- \ and \ B)}{0.0035}$$

Use the calculation box to find $P(- \ and \ B)$

```{r exercise1, exercise=TRUE}

```

```{r, exercise1-solution}
0.11*0.0035
```


```{r quiz5}
question("We know that P(N) + P(B) must equal 1. Why?",
  answer("Events N and B are independent - they are unrelated."),
  answer("Events N and B are complements - one or the other must occur, but not both.", correct=TRUE),
  answer("Events N and B are disjoint - they cannot occur simultaneously. There are also other possibilities.")
)
```

Similarly, we know that

$$P(+ \vert B) + P(- \vert B) = 1$$

This holds because these are mutually exclusive (or disjoint) events that cover all possibilities under the condition. For any woman who has breast cancer, a mammogram will either:

1. Detect the cancer, $+$.
2. Not detect the cancer, $-$.

Use the calculation box to find $P(+\vert B)$.

```{r exercise2, exercise=TRUE}

```

```{r, exercise2-solution}
1 - 0.11
```

## Bayes' Rule

Bayes' Rule is often used to "flip" a conditional probability statement. Given $P(A \vert B)$, we can use it to find $P(B \vert A)$ through this rule.

$$P(A \vert B) =  \frac{P(B \vert A)\times P(A)}{P(B)}$$
Suppose a woman over 40 in Canada gets a positive result on a mammogram. What is the probability that, given this positive result, she truly has breast cancer?

We know:

- $P(B) = 0.0035$
- $P(+) = 0.073$
- $P(+ \vert B) = 0.89$

The probability that given a positive test result a woman over 40 in Canada truly has breast cancer is

$$P(B \vert +) = \frac{P(+ \vert B)\times P(B)}{P(+)}= \frac{0.89\times 0.0035}{0.073}$$

Use the calculation box to find $P(B\vert +)$.

```{r exercise3, exercise=TRUE}

```

```{r, exercise3-solution}
0.89*0.0035/0.073
```


```{r quiz6}
question("Even with a positive mammogram result, the probability that a woman truly has breast cancer is low. Why?",
  answer("Breast cancer is a rare event in the population.", correct=TRUE),
  answer("Mammograms are unreliable."),
  answer("Diagnostic tests are always prone to error.")
)
```

Mammograms are often used as a "line of first defense" for detecting breast cancer. A positive mammogram result is often not a firm diagnosis - more tests will be done before diagnosing cancer.

## Practice

What proportion of breast cancer cases will go undetected each year in Canada? That is, given that a Canadian woman in her 40s has a negative mammogram screening, what is the probability she actually has cancer?

```{r quiz7}
question("Select the correct probability notation for this event.",
  answer("P(B|-)", correct=TRUE),
  answer("P(-|B)"),
  answer("P(B|+)"),
  answer("P(+|B)")
)
```

By Bayes' Rule, 

$$P(B \vert -) = \frac{P(- \vert B)\times P(B)}{P(-)}$$

Use the calculation box to find $P(-)$, the probability of a negative mammogram result.

```{r exercise4, exercise=TRUE}

```

```{r, exercise4-solution}
1 - 0.073
```

Use the calculation box to find $P(B \vert -)$, the probability that a woman with a negative mammogram result truly has breast cancer.

```{r exercise5, exercise=TRUE}

```

```{r, exercise5-solution}
0.11*0.0035/(1-0.073)
```

